---
title: "README.md"
author: "Wahsabii Neandertal"
date: "Thursday, February 19, 2015"
output: html_document 
---

***
* Task: create one R script called run_analysis.R that does the following. 
     + Part 1. Merges the training and the test sets to create one data set.
     + Part 2. Extracts only the measurements on the mean and standard deviation for each measurement. 
     + Part 3. Uses descriptive activity names to name the activities in the data set
     + Part 4. Appropriately labels the data set with descriptive variable names. 
     + Part 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

* In this project, deliver:
     + run_analysis.R : the R-code run on the data set
     + Tidy.txt : the clean data extracted from the original data using run_analysis.R
     + CodeBook.md : the CodeBook reference to the variables in Tidy.txt
     + README.md : the analysis of the code in run_analysis.R
     + analysis.html : the html version of README.md


### Use the following libraries.  
```{r message=FALSE}
library(dplyr)
library(data.table)
```

***
# Read the files into R
#
### Read metadata for descriptive names.  
```{r}
labelsActivity <- read.table("UCI HAR Dataset/activity_labels.txt", header = FALSE)  
namesFeatures  <- read.table("UCI HAR Dataset/features.txt")  
```

### Read Testing data from the uncompressed subdirectory in the project directory.  
```{r}
TestingSubject  <- read.table("UCI HAR Dataset/test/subject_test.txt", header = FALSE)  
TestingFeatures <- read.table("UCI HAR Dataset/test/X_test.txt", header = FALSE)  
TestingActivity <- read.table("UCI HAR Dataset/test/y_test.txt", header = FALSE)  
```

### Read Training data from the uncompressed subdirectory in the project directory.  
```{r}
TrainingSubject  <- read.table("UCI HAR Dataset/train/subject_train.txt", header = FALSE)  
TrainingFeatures <- read.table("UCI HAR Dataset/train/X_train.txt", header = FALSE)  
TrainingActivity <- read.table("UCI HAR Dataset/train/y_train.txt", header = FALSE)  
```

***
# Part 1: Merges the training and the test sets to create one data set.
```{r}
activity <- rbind(TestingActivity, TrainingActivity)  
features <- rbind(TestingFeatures, TrainingFeatures)  
subject  <- rbind(TestingSubject, TrainingSubject)  
```

### Descriptive feature names are added to the data columns after tranposing the column to a row vector.   
```{r}
colnames(features) <- t(namesFeatures$V2)  
```

### Make the final data product for analysis by column binding the data sets.  
```{r}
colnames(subject)  <- "Subject"  
colnames(activity) <- "Activity"  
mergedDataset      <- cbind(subject, activity, features)  
```

***
# Part 2: Extracts only the rows with mean and standard deviation statisitics.  
```{r}
statistics <- grep(".*Mean.*|.*Std.*", names(mergedDataset), ignore.case=TRUE)  
```

### Identify mean and standard deviation statistics.  
```{r}
statisticalColumns <- c(1, 2, statistics)  
```

### Subset the merged data set to only the statistics data set.  
```{r}
statisticalData <- mergedDataset[,statisticalColumns]  
```

***
# Part 3: Uses descriptive activity names to name the activities in the data set. 
```{r}
statisticalData$Activity <- as.character(statisticalData$Activity)  # convert to character
 
# Iterate through activity labels, subsitute the activity description from column 2 into the data
description <- 2  # second column has the description

for(row in seq_along(labelsActivity$V1)){  
      statisticalData$Activity[statisticalData$Activity == row] <- as.character(labelsActivity[row,description])  
}  
statisticalData$Activity <- as.factor(statisticalData$Activity)  # change to factor
```

***
# Part 4. Appropriately labels the data set with descriptive variable names.  

### Produce the abbreviated variable names.  
```{r}
names(statisticalData)  
```

### Unabbreviate all variable names.  
```{r}
names(statisticalData) <- gsub("Acc", "Accelerometer", names(statisticalData))  
names(statisticalData) <- gsub("Gyro", "Gyroscope", names(statisticalData))  
names(statisticalData) <- gsub("BodyBody", "Body", names(statisticalData))  
names(statisticalData) <- gsub("Mag", "Magnitude", names(statisticalData))  
names(statisticalData) <- gsub("^t", "Time", names(statisticalData))  
names(statisticalData) <- gsub("^f", "Frequency", names(statisticalData))  
names(statisticalData) <- gsub("tBody", "TimeBody", names(statisticalData))  
names(statisticalData) <- gsub("-mean()", "Mean", names(statisticalData), ignore.case = TRUE)  
names(statisticalData) <- gsub("-std()", "StandardDeviation", names(statisticalData), ignore.case = TRUE)  
names(statisticalData) <- gsub("-freq()", "Frequency", names(statisticalData), ignore.case = TRUE)  
names(statisticalData) <- gsub("angle", "Angle", names(statisticalData))  
names(statisticalData) <- gsub("gravity", "Gravity", names(statisticalData))  
names(statisticalData) <- gsub("X", "X-Coordinate", names(statisticalData))  
names(statisticalData) <- gsub("Y", "Y-Coordinate", names(statisticalData))  
names(statisticalData) <- gsub("Z", "Z-Coordinate", names(statisticalData))  
```

### Produce the final unabbreviated variable names. 
```{r}
names(statisticalData)  
```

***
# Part 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r}
answer <- data.table()
answer <- statisticalData %>% group_by(Subject, Activity) %>% summarise_each(funs(mean))
answer <- answer[order(answer$Subject, answer$Activity),]
write.table(answer, file = "Tidy.txt", row.names = FALSE)
```
#### References. 

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita and Xavier Parra. November 2013.

Website: http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones 

Data set: https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip 
